Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure has been driven by the 
needs of user communities to access large scale computing and data analysis infrastructure to 
support their research needs as part of their pan-European research collaborations. One of the most 
known examples is Worldwide Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, 
powered by the Particle Physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large Hadron
Collider particle accelerator. As a consequence a number of important projects have been run in Europe, supported by the European Commission funding of various research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
develop distributed solutions that provided a seamless and transparent access to those resources, available for the researchers.

Major developments were done throughout the years, mainly focused on providing new
functionalities, hindering the adoption of software engineering methodologies.
The software delivered was by then less tested, resulting in constant
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it become apparent that a balance was needed between new features addition
and the reliability of the software produced. As a consequence, a substantial part of the
software lifecycle would have to be devoted to deal with the quality of software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define the quality procedures, team
organization and pilot testbeds.

@@

In this paper, we introduce a set of European Commission (EC)-funded projects
that goes from 2001 up to now, where the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability. Finally,
Section \ref{sec:con} concludes.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data GridTwo & 2001--2004\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO-DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted
to develop the first European infrastructure for Grid computing. The project
brought together 21 academic and industry partners, from 15 different
countries. There were 3 application areas for requirement gathering;
Particle physics, Bioinformatics and Earth observation. The project delivered
their own software distribution, named EDG (EU DataGrid), strongly based on
Globus Middleware Services \cite{globus}. The development organization
consisted in 5 work packages coordinated by an Architecture Task Force,
supervising the overall design and technical consistency of the developments.

Due to having no previous experience in such widely
collaborative projects, a big effort was spent trying to devise solutions for several
challenges \cite{datagrid}, namely the 1) communication overhead, as a
result of the large geographical separation of the parties involved in the
development tasks, 2) the evolution of the requirements coming from the user
communities, and 3) the lack of a body of knowledge for academic software
engineering.

Agile methodologies were being introduced at the time
through the Agile manifesto \cite{agile-manifesto}, as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nonetheless, the project focused on the experiences and procedures
from open source software projects, such as Linux and Apache, trying to reach a
higher maturity level \cite{cmm}.

@@

The {\sl gLite} middleware \cite{glite} was the ultimate
official distribution of EGEE as of 2006, after two years of prototyping and
re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, that issued more than 10 thousand bugs, 1.7 thousand patches and
defined over 300 tasks tracked by the use of bug/task management tools.

The
source code was available at a private, centralized version control system.
The code was not validated by any Quality Assurance policy, but passed through a manual
certification procedure at the time of the release. This procedure tried to
improve the reliability of the software components by applying a set of
requirements [ref] to be fulfilled at each stage (integration, certification,
pre--production, production), carried out by different teams. However, starting in
EGEE--II, the project adopted the ETICS solution, aiming to develop
an automatic build system for Grid middleware.

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure 
has been driven by the needs of user communities to access large scale computing and 
data analysis infrastructure to support their research needs as part of their 
pan-European research collaborations. One of the most known examples is the Worldwide 
Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered 
by the Particle Physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. As a consequence a number of important projects 
have been run in Europe, supported by the European Commission funding of various 
research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
develop distributed solutions that provided a seamless and transparent access to those 
resources, available for the researchers.

Major developments were done throughout the years, mainly focused on providing new
functionalities, leaving secondly the adoption of software engineering methodologies.
The software delivered was then less tested, resulting in constant
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it become apparent that a balance was needed between new features addition
and the reliability of the software produced. As a consequence, a substantial part of the
software lifecycle would have to be devoted to deal with the quality of software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define the quality procedures, teams
organization and pilot testbeds.

@@

In this paper, we introduce a set of European Commission (EC)-funded projects
that goes from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability. Finally,
Section \ref{sec:con} concludes.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2004\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO-DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, Particle physics, Bioinformatics and Earth observation
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
%There were 3 application areas for requirement gathering; Particle physics, Bioinformatics and Earth observation. 
The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
based on Globus Middleware Services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

Due to having no previous experience in such widely
collaborative projects, a big effort was spent trying to devise solutions for several
challenges \cite{datagrid}, namely the 1) communication overhead, as a
result of the large geographical separation of the parties involved in the
development tasks, 2) the evolution of the requirements coming from the user
communities (50 use cases from the three scientific area), and 3) the lack of a body of knowledge for academic software
engineering.

Agile methodologies were being introduced at the time
through the Agile manifesto \cite{agile-manifesto}, as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nonetheless, the project focused on the experiences and procedures
used by different open source software projects, such as Linux and Apache, trying to reach a
higher maturity level \cite{cmm}.

@@

The {\sl gLite} middleware \cite{glite} was the ultimate
official software distribution of EGEE as of 2006, after two years of prototyping and
re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code was not validated by any Quality Assurance policy, but passed through a manual
certification procedure at the time of the release. This procedure tried to
improve the reliability of the software components by applying a set of
requirements [ref] to be fulfilled at each stage of the  development lifecycle (integration, certification,
pre--production, production), carried out by different teams. However, with 
EGEE--II, the developers started to use the the ETICS system [ref], solution aiming to develop
an automatic build system for Grid middleware.
